Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-30 Manuscript_Nater_etal.qmd
Minor edits (processing comments/feedback)

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

<!--# removed a few words/phrases that I thought were superfulous-->

@@

There is growing demand for biodiversity indicators from international unions, national governments, local management bodies, and corporate and industry actors. Such indicators should ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have high spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for infrastructure development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making [@christie2020]. Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This variation needs to be accounted for to develop successful and sustainable strategies for area use, harvest management, and species and biodiversity conservation [@williams2002].

@@

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program. This often limits the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence(-absence) or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates [@dickinson2010; @johnston2023]. The exception here are monitoring programmes that succeed in making use of a large number of volunteers that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>). In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) <!-- well, we do not really have a good citation (i.e. nothing published with a doi), but this is something we will release soon. The best option is to cite one of the papers where we have used the data (e.g. Bowler et al), or one of the GBIF data sets (where we describe the protocol in detail). --> and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF). As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for resident predators, such as the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988]. Across their distributional range, willow ptarmigan are iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is reported, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored to the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). As part of the sampling protocol, the field workers are also requested to report if they observed a selected set of mammalian species (small rodents, red foxes *Vulpes vulpes* and mountain hare *Lepus timidus)* while conducting the sampling*.* The small rodent observations from the line transect surveys will form the basis for our rodent co-variate. Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).

Notably, the program is not designed as a centralized national monitoring programme, but rather a collection of local and regional survey programs. All involved survey areas use a common field protocol and data collection model. In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided. However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous in space. In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and low in Western and Southern Norway.

@@

As part of the line transect sampling (see above), observers are also requested to report whether they have seen any small rodents while surveying a transect. For each transect survey, this information is recorded as 1's (small rodents spotted at least once) and 0's (no small rodents spotted). We aggregated this data into area- and year-specific rodent occupancy covariates by averaging the 0 and 1 reports for all transect surveys within a given area and year and subsequently z-standardizing values. We note that while we refer to the covariate as "rodent occupancy" throughout the manuscript, it can be interpreted as an index for rodent abundance. <!-- @ERLEND: I have added this section as it was missing. Anything you would like to add here? -->

<!-- I have added some text above - and think this is sufficient -->

@@

All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated. Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv". Supplementary figures (SF) are provided as .pdf files with captions in "SuppFigures_Captions.txt".

@@

The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`). Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`). This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf"). The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf"). Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C). <!-- ERLEND: we sometimes refer to rodent abundance, sometimes occupancy. I think what we have is most correctly refered to as an index of abundance. CHLOE: Good point about being consistent. I see that I have mostly used "occupancy" in text, figures, and code, and to me, that is also the most "precise" description. So I have standardized to primarily using "occupancy". Nonetheless, we should state that the way we treat occupancy becomes an abundance proxy. In fact, I noticed we were missing the description of rodent data. So added that, and made a mention there. Okay? ERLEND: Ok-->

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively. The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival and lower recruitment) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

We found increasing population trends over recent years in over half of the reporting districts, but population declines were also evident in some areas, particularly in the mountains in central Norway [@fig-lamA]. While we may speculate that recent population trends may be linked to changes in harvest regulations and/or climatic conditions, considering the whole time-series (2007-2021) illustrated that population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf"). At a national scale, recent analysis by @nilsen2020 indicates that the ptarmigan population in Norway increased between 2009 and 2020. In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends. Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018. Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf"), follwed by a low survival the year after (SF "TimeSeries_pSurv.pdf"). This often resulted in steep population declines towards the following year. The fact that these same years also match up with observed peaks in rodent abundance in many areas, together with the largely positive effects of rodent occupancy on recruitment estimated by our model (@fig-paramCorrs C), provides evidence for the Alternative Prey Hypothesis [APH; @hagen1952]. The APH stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020]. While @Nyström_etal_2006 suggested that gyrfalcons, which are specialized ptarmigan predators, do not respond to rodent populations or switch to alternative prey when ptarmigan populations are low, generalist predators, such as red foxes, are likely to shift from preying on ptarmigans to rodents when the latter become abundant [e.g. @Breisjøberget_etal_2018; @bowler2020]. Taking a spatial perspective, the highest latitude and highest altitude areas stood out once more, sporting the strongest effects of rodent occupancy (SF "betaR_Map.pdf"). This could be related to warmer areas generally having larger predator guilds, and consequently more generalists that are able to maintain relatively stable populations irrespective of small rodent abundance [@bowler2020].

Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. In previous work, both based on marked birds [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates. In general, it is found warmer and earlier springs result in earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall (as in our study). In @bowler2020, it was reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to the effect of rodent occupancy reported here. In practice, measures of the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data constitute relevant candidate covariates for future work alongside temperature.

>>>>

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

There is growing demand for biodiversity indicators from international unions, national governments, local management bodies, and corporate and industry actors. Indicators should ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have good spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for infrastructure development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for county- or municipality-level decision making [@christie2020]. Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This variation needs to be accounted for to develop successful and sustainable strategies for area use, harvest management, and species and biodiversity conservation [@williams2002].

@@

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and hinges on a sufficient degree of top-down control of the programme. This often limits the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence(-absence) or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates [@dickinson2010; @johnston2023]. The exception here are monitoring programmes that succeed in making use of a large number of volunteers that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. At the European level, the recently established initiative "European Observatory of Wildlife" is offering common field- and analyses protocols and aims to establish a network of "observation points" for monitoring wildlife populations at the European level (<https://wildlifeobservatory.org/>). In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF). As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for resident predators, such as the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988]. Across their distributional range, willow ptarmigan are an iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, fieldworkers use pointing dogs to locate the birds. A survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is reported, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored to the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).

Notably, the program is not designed as a centralized national monitoring programme, but rather a collection of local and regional survey programs. All involved survey areas use a common field protocol and data collection model. In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided. However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous across space. In general, sampling effort is higher in South-Eastern and Central Norway, intermediate in Northern Norway, and low in Western and Southern Norway.

@@

As part of the line transect sampling (see above), observers are also requested to report whether they have seen any small rodents while surveying a transect. For each transect survey, this information is recorded as 1's (small rodents spotted at least once) and 0's (no small rodents spotted). We aggregated this data into area- and year-specific rodent occupancy covariates by averaging the 0 and 1 reports for all transect surveys within a given area and year and subsequently z-standardizing values. We note that while we refer to the covariate as "rodent occupancy" throughout the manuscript, it can be interpreted as an index for rodent abundance.

@@

All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated. Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv". Supplementary figures (SFs) are provided as .pdf files with captions in "SuppFigures_Captions.txt"; all files are deposited on OSF (<https://osf.io/7326r/>).

@@

The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`). Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`). This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf"). The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf"). Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites ranging from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively (truncation distance = 200 m) . The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival and lower recruitment) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

We found increasing population trends over recent years in over half of the reporting districts, but population declines were also evident in some areas, particularly in the mountains in central Norway [@fig-lamA]. Predominantly increasing population trends are consistent with a recent national-scale analysis by @nilsen2020 which found an overall increase in the Norwegian ptarmigan population between 2009 and 2020. While we may speculate that recent population trends could be linked to changes in harvest regulations and/or climatic conditions, considering the whole time-series (2007-2021) illustrated that population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf"). In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends. Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018. Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf"), followed by a low survival the year after (SF "TimeSeries_pSurv.pdf"). This often resulted in steep population declines towards the following year. The fact that these same years also match up with observed peaks in rodent abundance in many areas, together with the largely positive effects of rodent occupancy on recruitment estimated by our model (@fig-paramCorrs C), provides evidence for the Alternative Prey Hypothesis [APH; @hagen1952]. The APH stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020]. While @Nyström_etal_2006 suggested that gyrfalcons, which are specialized ptarmigan predators, do not respond to rodent populations or switch to alternative prey when ptarmigan populations are low, generalist predators, such as red foxes, are likely to shift from preying on ptarmigans to rodents when the latter become abundant [e.g. @Breisjøberget_etal_2018; @bowler2020]. Taking a spatial perspective, the highest latitude and highest altitude areas stood out once more, sporting the strongest effects of rodent occupancy (SF "betaR_Map.pdf"). This could be related to warmer areas generally having larger predator guilds, and consequently more generalists that are able to maintain relatively stable populations irrespective of small rodent abundance [@bowler2020].

Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates. In general, warmer and earlier springs seem to favour earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall. @bowler2020 further reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to what we found for the effect of rodent occupancy here. In practice, measures representing spring conditions, such as the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data, thus constitute relevant candidate covariates for future work alongside temperature.
